---
title: "plyinteractions"
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('plyinteractions')`"
vignette: >
  %\VignetteIndexEntry{plyinteractions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Pre-load libraries
library(GenomicRanges)
library(InteractionSet)
library(plyinteractions)

## Bib setup
library("RefManageR")
bib <- c(
    R = citation(),
    InteractionSet = citation("InteractionSet")[1],
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    dplyr = citation("dplyr")[1],
    rlang = citation("rlang")[1],
    plyranges = citation("plyranges")[1],
    plyinteractions = citation("plyinteractions")[1]
)
```

The `r Biocpkg("plyinteractions")` package
introduces tidy methods for the `GInteractions` class defined in the 
`r Biocpkg("InteractionSet")` package `r Citep(bib[["InteractionSet"]])`. 

# Install `plyinteractions`

`r Biocpkg("plyinteractions")` is a `R` package available via 
[Bioconductor](http://bioconductor.org). 
You can install `r Biocpkg("plyinteractions")` by using the following 
commands in your `R` session:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("plyinteractions")
```

`GInteractions` are objects describing interactions between pairs of genomic 
ranges: 

```{r}
library(GenomicRanges)
library(InteractionSet)
anchor1 <- GRanges("chr1:10-20:+")
anchor2 <- GRanges("chr1:50-60:-")
gi <- GInteractions(anchor1, anchor2)

gi
```

# Importing genomic interactions in R

`r Biocpkg("plyinteractions")` provides a consistent interface for importing 
genomic interactions from `pairs` and `bedpe` files into GInteractions in R, 
following grammar of tidy data manipulation defined in the 
`r CRANpkg("tidyverse")` ecosystem. 

Tidy genomic data maniuplation implies that we first parse genomic 
files stored on disk as tabular data frames. 

```{r "start", message=FALSE}
bedpe_file <- system.file("tests", "test.bedpe", package = "rtracklayer")
bedpe_df <- read.delim(bedpe_file, header = FALSE, sep = '\t')

bedpe_df
```

Genomic interactions in tabular format are not easy to manipulate.
We can easily parse a `data.frame` into a `GInteractions` object using 
the `as_ginteractions` function. 

```{r}
gi <- bedpe_df |> 
    as_ginteractions(
        seqnames1 = V1, start1 = V2, end1 = V3, strand1 = V9, 
        seqnames2 = V4, start2 = V5, end2 = V6, strand2 = V10, 
        starts.in.df.are.0based = TRUE
    )

gi
```

The columns containing information for core fields of the future `GInteractions` 
object (e.g. `seqnames1`, `strand2`, ...) can be specified using the 
`key = value` (supported by quasiquotation). 

The reverse operation to coerce `GInteractions` back to a tabular form is 
also possible using the `as_tibble` function from the `tibble` package: 

```{r}
tibble::as_tibble(gi)
```

# Manipulating `GInteractions`

`r Biocpkg("plyinteractions")` provides a set of verbs for developing 
analysis pipelines based on `GInteractions` objects that represent genomic 
interactions. 

## Mutating columns 

`mutate` supports accessing other existing columns:

```{r}
gi |> mutate(cis = seqnames1 == seqnames2)

gi |> mutate(both_chr = paste(seqnames1, seqnames2, sep = "_"))

gi |> mutate(start1 = 1)
```

## Grouping columns 

`group_by` supports accessing both core and metadata columns:

```{r}
gi |> group_by(seqnames2)

gi |> group_by(cis = seqnames1 == seqnames2)

gi |> group_by(seqnames2, cis = seqnames1 == seqnames2)
```

## Summarizing columns 

Summarizing grouped GInteractions can be extremely powerful. 

```{r}
pairs_file <- HiContactsData::HiContactsData('yeast_wt', 'pairs.gz')
pairs_df <- read.delim(pairs_file, sep = "\t", header = FALSE, comment.char = "#")
head(pairs_df)
pairs <- as_ginteractions(pairs_df, 
    seqnames1 = V2, start1 = V3, strand1 = V6, 
    seqnames2 = V4, start2 = V5, strand2 = V7, 
    width1 = 1, width2 = 1, 
    keep.extra.columns = FALSE
)
pairs

pairs |> group_by(same_strand = strand1 == strand2) |> 
    summarize(
        neg_strand = sum(strand1 == "-"), 
        pos_strand = sum(strand1 == "+")
    )
```

## Filtering columns 

`filter` supports logical expressions:

```{r}
gi |> filter(seqnames1 == 'chr11')

gi |> filter(start1 >= 1e8)

gi |> filter(seqnames1 == seqnames2)
```

## Selecting columns 

`select` supports [`<tidy-select>` arguments](https://dplyr.tidyverse.org/reference/dplyr_tidy_select.html):

```{r}
gi |> select(V7)

gi |> select(contains("8"))

gi |> select(starts_with("V"))
```

Note that core fields (e.g. `seqnames1`, `strand2`, ...) cannot be retrieved 
using this approach, only metadata columns are parsed. 
Selecting a subset of core fields from a `GInteractions` would lead to 
loss of required information (the other non-selected core fields). 

```{r}
## This does not restrict to `seqnames1` and `seqnames2` columns. 
gi |> select(starts_with('seq')) 
```

Forcing selection of core fields is still possible when using the `.drop_ranges`
argument of `select()`. This results in the coercion of the selected columns 
into a `DataFrame`. 

```{r}
## This selects `seqnames1` and `seqnames2` columns but converts the output
## into a `DataFrame`.
gi |> select(starts_with('seq'), .drop_ranges = TRUE) 
```

## Slicing rows 

```{r}
gi |> slice(1, 2)
gi |> slice(-3)
```

# Citing `plyinteractions`

We hope that `r Biocpkg("plyinteractions")` will be useful for your research. 
Please use the following information to cite the package and the overall 
approach. Thank you!

```{r "citation"}
## Citation info
citation("plyinteractions")
```

# Acknowledgments

The `r Biocpkg("plyinteractions")` package 
introduces tidy methods for the `GInteractions` class defined in the 
`r Biocpkg("InteractionSet")` package `r Citep(bib[["InteractionSet"]])`. 

The `r Biocpkg("plyinteractions")` package 
follows tidy principles defined for tabular data and genomic ranges:

* `r CRANpkg("dplyr")` `r Citep(bib[["dplyr"]])`
* `r CRANpkg("rlang")` `r Citep(bib[["rlang"]])`
* `r Biocpkg("plyranges")` `r Citep(bib[["plyranges"]])`

The `r Biocpkg("plyinteractions")` package `r Citep(bib[["plyinteractions"]])`
was written using the following resources:

* R `r Citep(bib[["R"]])`
* `r CRANpkg("devtools")`
* `r CRANpkg("usethis")`

Supporting documentation was generated using the following resources: 

* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`

# Reproducibility 

`R` session information:

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```

# Bibliography

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
